---
title: "STAT 440/540 Milestone 3: Multiple Regression & Refinement"
author: "Group 03: Marilla Cabrigas Blatnick, Kaylie Doyle, Carly Salazar"
date: "2025-10-30"
format:
  pdf:
    toc: true
    number-sections: true
    geometry: margin=1in
execute:
  echo: true
  warning: false
  message: false
editor: visual
---

# Introduction

-   Recap of EDA and initial insights guiding model building.

## Package Setup

```{r}
#| label: pkg-setup
if (!requireNamespace("pacman", quietly = TRUE)) install.packages("pacman")
pacman::p_load(tidyverse, broom, car, performance, interactions, splines, mgcv,
               kableExtra, ggplot2, patchwork)
set.seed(440540)
```

## Load Data

```{r}
#| label: load-data

# Load the data (keep original column names)
dat <- readr::read_csv("../data/raw/student_lifestyle_dataset.csv")

# Convert Stress_Level from text to numbers
dat <- dat |>
  dplyr::mutate(
    Stress_Level = dplyr::case_when(
      Stress_Level == "Low" ~ 1,
      Stress_Level == "Moderate" ~ 2,
      Stress_Level == "High" ~ 3
    )
  )
```

## Candidate Models

We consider linear, interaction, polynomial, and spline specifications.

```{r}
#| label: fit-candidates

# MODEL 1: Linear (Main Effects Only)
m_lin <- lm(GPA ~ Study_Hours_Per_Day + Physical_Activity_Hours_Per_Day + 
            Social_Hours_Per_Day + Sleep_Hours_Per_Day + Stress_Level, 
            data = dat)

# MODEL 2: Interaction - Study Hours × Stress Level
m_int <- lm(GPA ~ Study_Hours_Per_Day * Stress_Level + 
            Physical_Activity_Hours_Per_Day + Social_Hours_Per_Day + 
            Sleep_Hours_Per_Day, 
            data = dat)

# MODEL 3: Polynomial - Quadratic Sleep Effect
m_poly <- lm(GPA ~ Study_Hours_Per_Day + Physical_Activity_Hours_Per_Day + 
             Social_Hours_Per_Day + Sleep_Hours_Per_Day + I(Sleep_Hours_Per_Day^2) + 
             Stress_Level, 
             data = dat)

# MODEL 4: Spline - Flexible nonlinear effect for study hours
library(splines)
m_spline <- lm(GPA ~ bs(Study_Hours_Per_Day, df=4) + Physical_Activity_Hours_Per_Day + 
               Social_Hours_Per_Day + Sleep_Hours_Per_Day + Stress_Level, 
               data = dat)
```

## Model Summaries

```{r}
#| label: tbl-candidates
##| tbl-cap: "Model fit summaries."
purrr::map_df(
  list(lin=m_lin, int=m_int, poly=m_poly, spline=m_spline),
  ~broom::glance(.x), .id="model"
) |>
  dplyr::select(model, r.squared, adj.r.squared, AIC, BIC, sigma) |>
  knitr::kable(digits=3)
```

## Coefficient Interpretation

```{r}
#| label: tbl-coefs
##| tbl-cap: "Coefficient estimates for interaction model."
broom::tidy(m_int, conf.int = TRUE) |>
  knitr::kable(digits=3)
```

## Diagnostics

```{r}
##| fig-cap: "Residual diagnostics for interaction model."
#| fig-width: 12
#| fig-height: 10

# Display each diagnostic plot separately
par(mfrow = c(1, 1))  # Single plot at a time
plot(m_int, which = 1)  # Residuals vs Fitted
plot(m_int, which = 2)  # Q-Q plot
plot(m_int, which = 3)  # Scale-Location
plot(m_int, which = 5)  # Residuals vs Leverage
```

```{r}
#| label: vif
car::vif(m_int)
```

## Interaction Plot

```{r}
#| label: fig-interactions
##| fig-cap: "Interaction effect visualization."
#| fig-width: 10
#| fig-height: 6

# Create prediction data
pred_data <- expand.grid(
  Study_Hours_Per_Day = seq(5, 10, by = 0.5),
  Stress_Level = c(1, 2, 3),
  Physical_Activity_Hours_Per_Day = mean(dat$Physical_Activity_Hours_Per_Day),
  Social_Hours_Per_Day = mean(dat$Social_Hours_Per_Day),
  Sleep_Hours_Per_Day = mean(dat$Sleep_Hours_Per_Day)
)

# Get predictions
pred_data$predicted_gpa <- predict(m_int, newdata = pred_data)

# Plot
library(ggplot2)
ggplot(pred_data, aes(x = Study_Hours_Per_Day, y = predicted_gpa, 
                      color = factor(Stress_Level), group = factor(Stress_Level))) +
  geom_line(size = 1.2) +
  labs(
    title = "How Stress Moderates the Effect of Study Hours on GPA",
    x = "Study Hours per Day",
    y = "Predicted GPA",
    color = "Stress Level"
  ) +
  scale_color_manual(
    values = c("1" = "green", "2" = "orange", "3" = "red"),
    labels = c("1" = "Low", "2" = "Moderate", "3" = "High")
  ) +
  theme_minimal()
```

## Model Selection Rationale

-   Compare metrics and diagnostics; justify the preferred model.
-   Note limitations and trade-offs (bias–variance, interpretability).

## Reproducibility Snapshot

```{r}
#| label: session-info
sessionInfo()
```
