---
title: "STAT 440/540 Milestone 2: EDA & Initial Models"
author: "Group 03: Marilla Cabrigas Blatnick, Kaylie Doyle, Carly Salazar"
date: "2025-09-25"
format:
  pdf:
    toc: true
    number-sections: true
    geometry: margin=1in
execute:
  echo: true
  warning: false
  message: false
editor: visual
---

# Overview

-   Briefly restate question, dataset, and key variables.

The purpose of the this project is to explore the relationship that mental health and other life style factors that can affect a college student's GPA. 
This dataset gathers data on GPA, mental health factors, student habits, and other life style factors. 
The key variables in this data set is the repsonse variable *GPA* and other important variables such as sleep and [something else].

## Package Setup

```{r}
#| label: pkg-setup
if (!requireNamespace("pacman", quietly = TRUE)) install.packages("pacman")
pacman::p_load(tidyverse, janitor, skimr, GGally, broom, patchwork, knitr, kableExtra)
set.seed(440540)
```

## Data Import and Cleaning

```{r}
#| label: data-import
raw <- readr::read_csv("../data/raw/student_lifestyle_dataset.csv") |> janitor::clean_names()

# raw <- readr::read_csv("../data/raw/example_stub.csv") |> janitor::clean_names()

summary(raw) 

# dat <- raw 

dat <- raw |>
  dplyr::select(-student_id)



# Code originally given
  # mutate(
  #   x3 = factor(x3)
  # ) |>
  # tidyr::drop_na(y)     
```

## Descriptive Statistics

```{r}
#| label: tbl-summary
#| tbl-cap: "Summary statistics for numeric variables."

summary_data <- dat |>
  dplyr::select(where(is.numeric)) |>
  skimr::skim()

# First half of kable
summary_data |>
  dplyr::select(
    skim_type,
    skim_variable,
    n_missing,
    complete_rate,
    numeric.mean,
    numeric.sd,
  ) |>
  knitr::kable()

# Second half of kable
summary_data |>
  dplyr::select(
    skim_variable,
    numeric.p0,
    numeric.p25,
    numeric.p50,
    numeric.p75,
    numeric.p100
  ) |>
  knitr::kable()


# THE original CODE GIVEN
# dat |>
#   dplyr::select(where(is.numeric)) |>
#   skimr::skim() |>
#   knitr::kable()

```

## Exploratory Visualizations

```{r}
#| label: fig-pairs
#| fig-cap: "Pairs plot for numeric variables."

# ---- From old data set
# gpa_dat <- dat |>
#   dplyr::select(academic_performance_gpa, study_hours_per_week, financial_stress)

# gpa_dat2 <- dat |>
#   dplyr::select(academic_performance_gpa, mental_stress_level, diet_quality, cognitive_distortions, substance_use)

# ----

GGally::ggpairs(dplyr::select(dat, where(is.numeric)))
```

```{r}
#| label: fig-y-vs-x
#| fig-cap: "Outcome vs. selected predictors with loess smooth."
dat |>
  dplyr::select(where(is.numeric)) |>
  tidyr::pivot_longer(cols = -c(gpa), names_to = "var", values_to = "val") |>
  ggplot(aes(val, gpa)) +
  geom_point(alpha = 0.1) +
  geom_smooth(method = "loess", se = TRUE) +
  facet_wrap(~ var, scales = "free_x")

  # dat |>
  # dplyr::select(where(is.numeric)) |>
  # tidyr::pivot_longer(cols = -c(y), names_to = "var", values_to = "val") |>
  # ggplot(aes(val, y)) +
  # geom_point(alpha = 0.5) +
  # geom_smooth(method = "loess", se = TRUE) +
  # facet_wrap(~ var, scales = "free_x")
```

## Initial Simple Regressions

```{r}
#| label: simple-models
cand_x <- c("study_hours_per_day","extracurricular_hours_per_day","sleep_hours_per_day", "social_hours_per_day", "physical_activity_hours_per_day")
fits <- lapply(cand_x, \(v) lm(reformulate(v, response = "gpa"), data = dat))
models <- tibble::tibble(
  predictor = cand_x,
  glance = purrr::map(fits, broom::glance),
  tidy = purrr::map(fits, broom::tidy)
)

dplyr::bind_rows(purrr::map2(models$predictor, models$glance, \(v,g) dplyr::mutate(g, predictor=v))) |>
  dplyr::select(predictor, r.squared, adj.r.squared, AIC, BIC, p.value) |>
  knitr::kable(digits = 3, caption = "Simple regression comparisons")

# cand_x <- c("x1","x2","x3")
# fits <- lapply(cand_x, \(v) lm(reformulate(v, response = "y"), data = dat))
# models <- tibble::tibble(
#   predictor = cand_x,
#   glance = purrr::map(fits, broom::glance),
#   tidy = purrr::map(fits, broom::tidy)
# )

# dplyr::bind_rows(purrr::map2(models$predictor, models$glance, \(v,g) dplyr::mutate(g, predictor=v))) |>
#   dplyr::select(predictor, r.squared, adj.r.squared, AIC, BIC, p.value) |>
#   knitr::kable(digits = 3, caption = "Simple regression comparisons")
```

## Early Findings and Next Steps

-   What looks promising? Any evidence of nonlinearity or interactions to explore?

## Reproducibility Snapshot

```{r}
#| label: session-info
sessionInfo()
```