---
title: "STAT 440/540 Project Proposal"
author: "Group XX: Name1, Name2, Name3"
date: "2025-09-04"
format:
  pdf:
    toc: true
    number-sections: false
    geometry: margin=1in
execute:
  echo: true
  warning: false
  message: false
editor: visual
---

# Project Proposal

## Group and Communication Plan

-   Members and emails
-   Collaboration tools (e.g., GitHub, Google Drive) and cadence
-   Roles (analysis, writing, QA)

## Research Question

-   Primary question (predictive or explanatory)
-   Target outcome (Y) (continuous)
-   Why this matters (context, stakeholders)

## Dataset Summary

-   Source and link
-   Observations (n) and variables (p)
-   Candidate predictors (X) (quantitative/categorical)
-   Licensing/usage constraints

## Data Dictionary (Concise)

-   Variable \| Type \| Description \| Units \| Expected role (Y / X)
-   …

## Suitability for Regression

-   Linearity/plausibility, expected signal, anticipated interactions
-   Potential data issues (missingness, outliers, measurement error)

## Risks and Mitigations

-   Data access, cleaning complexity, time constraints → proposed mitigations

## Ethical Use and Reproducibility

-   Privacy, bias, fairness considerations
-   Plan for reproducible workflows (versioning, seeds, session info)

## Planned Analytic Path

-   EDA → simple regressions → candidate multiple models → diagnostics → refinement

## Package Setup

```{r}
#| label: pkg-setup
#| include: true
if (!requireNamespace("pacman", quietly = TRUE)) install.packages("pacman")
pacman::p_load(tidyverse, janitor, broom, knitr, kableExtra, ggplot2)
set.seed(440540)
```

## Reproducibility Snapshot

```{r}
#| label: session-info
sessionInfo()
```
